---
title: "--- Day 4: Passport Processing ---"
author: Fleur Kelpin
date: Dec 3, 2020
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
input <- readr::read_file("day04.txt") %>%
  str_split("\n\n") %>%
  unlist()
input <- tibble(line = input)
```

```{r}
passports <- input %>%
  extract("line",
    into = "byr", regex = "byr:([^\\s]+)", remove = FALSE, convert = TRUE
  ) %>%
  extract("line",
    into = "iyr", regex = "iyr:([^\\s]+)", remove = FALSE, convert = TRUE
  ) %>%
  extract("line",
    into = "eyr", regex = "eyr:([^\\s]+)", remove = FALSE, convert = TRUE
  ) %>%
  extract("line", into = "hgt", regex = "hgt:([^\\s]+)", remove = FALSE) %>%
  extract("line", into = "hcl", regex = "hcl:([^\\s]+)", remove = FALSE) %>%
  extract("line", into = "ecl", regex = "ecl:([^\\s]+)", remove = FALSE) %>%
  extract("line", into = "pid", regex = "pid:([^\\s]+)", remove = FALSE) %>%
  extract("line", into = "cid", regex = "cid:([^\\s]+)")
passports
```
# Part 1
Count the number of valid passports - those that have all required fields.
Treat cid as optional. **In your batch file, how many passports are valid?**

```{r}
drop_na(passports, -cid) %>%
  count()
```

# Part 2

You can continue to ignore the cid field, but each other field has strict rules
about what values are valid for automatic validation:

* byr (Birth Year) - four digits; at least 1920 and at most 2002.
* iyr (Issue Year) - four digits; at least 2010 and at most 2020.
* eyr (Expiration Year) - four digits; at least 2020 and at most 2030.
* hgt (Height) - a number followed by either cm or in:
  * If cm, the number must be at least 150 and at most 193.
  * If in, the number must be at least 59 and at most 76.
* hcl (Hair Color) - a # followed by exactly six characters 0-9 or a-f.
* ecl (Eye Color) - exactly one of: amb blu brn gry grn hzl oth.
* pid (Passport ID) - a nine-digit number, including leading zeroes.
* cid (Country ID) - ignored, missing or not.

Count the number of **valid** passports - those that have all required fields
**and valid values**. Continue to treat cid as optional. **In your batch file,
how many passports are valid?**

```{r}
passports %>%
  extract(hgt,
    into = c("hgt", "hgt_dim"), regex = "^([\\d]+)(cm|in)$", convert = TRUE
  ) %>%
  mutate(hgt_dim = as_factor(hgt_dim)) %>%
  drop_na(-cid) %>%
  filter(
    byr %>% between(1920, 2002),
    iyr %>% between(2010, 2020),
    eyr %>% between(2020, 2030),
    hgt_dim == "cm" & between(hgt, 150, 193) |
      hgt_dim == "in" & between(hgt, 59, 76),
    hcl %>% str_detect("^#[0-9a-f]{6}$"),
    ecl %>% str_detect("^amb|blu|brn|gry|grn|hzl|oth$"),
    pid %>% str_detect("^[0-9]{9}$")
  ) %>%
  count()
```
