---
title: "--- Day 2: Password Philosophy ---"
author: Fleur Kelpin
date: Dec 2, 2020
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
input <- read_csv("day02.txt",
  col_names = "line",
  col_types = cols(line = col_character())
) %>%
  extract(line,
    "(\\d+)-(\\d+) (\\w): (\\w+)",
    into = c("min", "max", "char", "pwd")
  ) %>%
  mutate(min = strtoi(min, 10), max = strtoi(max, 10))
```

# Part 1
Each line gives the password policy and then the password. The password policy
indicates the lowest and highest number of times a given letter must appear for
the password to be valid. For example, 1-3 a means that the password must
contain a at least 1 time and at most 3 times.

**How many passwords are valid** according to their policies?

```{r}
input %>%
  mutate(occurs = str_count(pwd, char)) %>%
  filter(occurs >= min) %>%
  filter(occurs <= max) %>%
  count()
```

# Part 2
Each policy actually describes two positions in the password, where 1 means the
first character, 2 means the second character, and so on. (Be careful; Toboggan
Corporate Policies have no concept of "index zero"!) **Exactly one of these
positions** must contain the given letter. Other occurrences of the letter are
irrelevant for the purposes of policy enforcement.

```{r}
input %>%
  mutate(first = str_sub(pwd, min, min)) %>%
  mutate(second = str_sub(pwd, max, max)) %>%
  filter(char == first | char == second) %>%
  filter(first != second) %>%
  count()
```
