---
title: "--- Day 16: Ticket Translation ---"
author: Fleur Kelpin
date: Dec 16, 2020
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
fields <- tibble(line = readLines("day16-fields.txt")) %>%
  extract(line,
    into = c("field", "from1", "to1", "from2", "to2"),
    "([^:]*): ([0-9]+)-([0-9]+) or ([0-9]+)-([0-9]+)",
    convert = T
  )
fields
```
```{r}
your_ticket <-
  "137,173,167,139,73,67,61,179,103,113,163,71,97,101,109,59,131,127,107,53"
nearby_tickets <- tibble(values = readLines("day16-nearby.txt"))
nearby_tickets
```
# Part 1
Start by determining which tickets are completely invalid; these are tickets
that contain values which aren't valid for any field. Ignore your ticket for
now.

Adding together all of the invalid values produces your ticket scanning error
rate.

Consider the validity of the nearby tickets you scanned. What is your ticket
scanning error rate?

```{r}
valid_value <- function(value) {
  fields %>%
    rowwise() %>%
    filter(between(value, from1, to1) | between(value, from2, to2)) %>%
    nrow() > 0
}

ticket_scanning_error_rate <- function(values) {
  invalid <- tibble(value = values) %>%
    separate_rows(value, sep = ",", convert = T) %>%
    rowwise() %>%
    filter(!valid_value(value))
  sum(invalid$value)
}

part1 <- nearby_tickets %>%
  rowwise() %>%
  mutate(error = ticket_scanning_error_rate(values))
sum(part1$error)
```
# Part 2

Now that you've identified which tickets contain invalid values, discard those
tickets entirely. Use the remaining valid tickets to determine which field is
which.
```{r}
valid_tickets <- part1 %>%
  filter(error == 0) %>%
  {
    c(.$values, your_ticket)
  }
tibble(valid_tickets)
```

Using the valid ranges for each field, determine what order the fields appear on
the tickets. The order is consistent between all tickets: if seat is the third
field, it is the third field on every ticket, including your ticket.
```{r}
valid_values <- valid_tickets %>%
  map(function(value) {
    str_split(value, ",")
  }) %>%
  unlist() %>%
  as.integer() %>%
  matrix(nrow = length(valid_tickets), byrow = T)
```
We're drowning in numbers here, let's try and summarize them in a usable way.
```{r}
# index_options[[i]] is true if this field can be column with index i in the
# valid_values list
index_options <- function(from1, to1, from2, to2) {
  valid_values %>%
    map_lgl(~ between(., from1, to1) | between(., from2, to2)) %>%
    matrix(nrow = length(valid_tickets), byrow = F) %>%
    colSums() %>%
    map_lgl(~ . == length(valid_tickets))
}
```
Matrix `x[i,j]` indicates if field `i` can have index `j` on the ticket.
```{r}
x <- fields %>%
  rowwise() %>%
  mutate(index_options = list(index_options(from1, to1, from2, to2)))
x <- matrix(unlist(x$index_options), ncol = nrow(fields), byrow = T)
```
Now if a row has only one TRUE value, we've found its index on the ticket.
Fill in the index, and set the row and column to false, cause we know that
other fields must match other indices. Rinse and repeat.
```{r}
fields$index = NA_integer_
while (any(x)) {
  row <- rowSums(x) %>%
    map_lgl(~ . == 1) %>%
    which()
  row <- row[[1]]
  col <- which(x[row, ])
  col <- col[[1]]
  fields$index[row] <- col
  x[row, ] <- F
  x[, col] <- F
}
```

Once you work out which field is which, look for the six fields on your ticket
that start with the word departure. What do you get if you multiply those six
values together?
```{r}
options(digits = 22)
fields %>%
  filter(str_detect(field, "departure")) %>%
  rowwise() %>%
  mutate(value = as.integer(unlist(str_split(your_ticket, ","))[[index]])) %>%
  ungroup() %>%
  summarize(prod(value))
```
