---
title: "--- Day 14: Docking Data ---"
author: Fleur Kelpin
date: Dec 14, 2020
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
input <- tibble(line = readLines("day14.txt")) %>%
  extract(line, "mask", "mask = ([01X]{36})", remove = FALSE) %>%
  extract(line, c("address", "value"), "mem\\[(\\d+)\\] = (\\d+)", convert = TRUE)
input
```
# Part 1

The initialization program (your puzzle input) can either update the bitmask or
write a value to memory. Values and memory addresses are both 36-bit unsigned
integers. For example, ignoring bitmasks for a moment, a line like `mem[8] = 11`
would write the value 11 to memory address 8.

The bitmask is always given as a string of 36 bits, written with the most
significant bit (representing 2^35) on the left and the least significant bit
(2^0, that is, the 1s bit) on the right. The current bitmask is applied to
values immediately before they are written to memory: a 0 or 1 overwrites the
corresponding bit in the value, while an X leaves the bit in the value
unchanged.

Execute the initialization program. What is the sum of all values left in memory
after it completes?

```{r}
i2b <- function(value) {
  result <- logical(36)
  index <- 36
  while (value > 0) {
    result[index] <- value %% 2 == 1
    value <- value %/% 2
    index <- index - 1
  }
  result
}

b2d <- function(bits) {
  result <- 0
  for (index in 1:36) {
    if (bits[[index]]) {
      result <- result + 2**(36 - index)
    }
  }
  result
}

apply_mask <- function(value, mask) {
  bits <- i2b(value)
  mask <- unlist(str_split(mask, ""))
  masked_bits <-
    map_lgl(1:36, function(index) {
      switch(mask[index],
        "1" = TRUE,
        "0" = FALSE,
        "X" = bits[index]
      )
    })
  b2d(masked_bits)
}
```
```{r}
mem <- double()

for (i in 1:nrow(input)) {
  if (!is.na(input$mask[[i]])) {
    mask <- input$mask[[i]]
  } else {
    mem[input$address[[i]]] <- apply_mask(input$value[[i]], mask)
  }
}

options(digits = 22)
sum(mem, na.rm = TRUE)
```

# Part 2

For some reason, the sea port's computer system still can't communicate with
your ferry's docking program. It must be using version 2 of the decoder chip!

A version 2 decoder chip doesn't modify the values being written at all.
Instead, it acts as a memory address decoder. Immediately before a value is
written to memory, each bit in the bitmask modifies the corresponding bit of the
destination memory address in the following way:

* If the bitmask bit is 0, the corresponding memory address bit is unchanged.
* If the bitmask bit is 1, the corresponding memory address bit is overwritten
  with 1.
* If the bitmask bit is X, the corresponding memory address bit is floating.

A floating bit is not connected to anything and instead fluctuates
unpredictably. In practice, this means the floating bits will take on all
possible values, potentially causing many memory addresses to be written all at
once!

```{r}
library(hash)
expand_masks <- function(mask) {
  if (str_detect(mask, "X")) {
    c(
      expand_masks(str_replace(mask, "X", "0")),
      expand_masks(str_replace(mask, "X", "1"))
    )
  } else {
    b2d(unlist(str_split(mask, "")) == "1")
  }
}
apply_mask_v2 <- function(value, mask) {
  bits <- i2b(value)
  mask <- unlist(str_split(mask, ""))
  applied <- map_chr(1:36, function(index) {
    switch(mask[index],
      "1" = "1",
      "0" = if (bits[[index]]) "1" else "0",
      "X" = "X"
    )
  })
  expand_masks(str_c(applied, collapse = ""))
}
apply_mask_v2(42, "000000000000000000000000000000X1001X")
```
```{r}
mem <- hash()
for (i in 1:nrow(input)) {
  if (!is.na(input$mask[[i]])) {
    mask <- input$mask[[i]]
  } else {
    addresses <- apply_mask_v2(input$address[[i]], mask)
    for (j in 1:length(addresses)) {
      mem[addresses[[j]]] <- input$value[[i]]
    }
  }
}
sum(values(mem))
```
