---
title: "--- Day 11: Seating System ---"
author: Fleur Kelpin
date: Dec 11, 2020
output: github_document
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(Matrix)
input <- read_csv(file = "day11.txt", col_names = c("seat"))
t <- input %>%
  separate_rows(seat, sep = "", convert = TRUE) %>%
  drop_na()
seats <- matrix(data = t$seat, nrow = nrow(input), byrow = TRUE)
dim(seats)
```

# Part 1

Now, you just need to model the people who will be arriving shortly.
Fortunately, people are entirely predictable and always follow a simple set of
rules. All decisions are based on the number of occupied seats adjacent to a
given seat (one of the eight positions immediately up, down, left, right, or
diagonal from the seat). The following rules are applied to every seat
simultaneously:

* If a seat is empty (L) and there are no occupied seats adjacent to it,
  the seat becomes occupied.
* If a seat is occupied (#) and four or more seats adjacent to it are also
  occupied, the seat becomes empty.
* Otherwise, the seat's state does not change.

[Game of life](
https://win-vector.com/2018/10/28/conways-game-of-life-in-r-or-on-the-importance-of-vectorizing-your-r-code/) with a mask. R is supposed to be good with matrices, so this should be a breeze.

```{r}
life_step <- function(seats, d) {
  # form the neighboring sums
  nrow <- dim(d)[[1]]
  ncol <- dim(d)[[2]]
  d_eu <- rbind(d[-1, , drop = FALSE], 0)
  d_ed <- rbind(0, d[-nrow, , drop = FALSE])
  d_le <- cbind(d[, -1, drop = FALSE], 0)
  d_re <- cbind(0, d[, -ncol, drop = FALSE])
  d_lu <- cbind(d_eu[, -1, drop = FALSE], 0)
  d_ru <- cbind(0, d_eu[, -ncol, drop = FALSE])
  d_ld <- cbind(d_ed[, -1, drop = FALSE], 0)
  d_rd <- cbind(0, d_ed[, -ncol, drop = FALSE])
  pop <- d_eu + d_ed + d_le + d_re + d_lu + d_ru + d_ld + d_rd
  d <- (!d & pop == 0) | (d & pop < 4)
  d & seats
}
state <- seats & FALSE
next_state <- life_step(seats, state)
while (any(next_state != state)) {
  state <- next_state
  next_state <- life_step(seats, state)
}
sum(state)
```

# Part 2
As soon as people start to arrive, you realize your mistake. People don't just
care about adjacent seats - they care about the first seat they can see in each
of those eight directions!

Now, instead of considering just the eight immediately adjacent seats, consider
the first seat in each of those eight directions.

I realise my mistake: to do advent of code in R!

R likes vectors, so let's number the seats, then the state becomes a boolean
vector.
```{r}
seat_coords <- as_tibble(which(seats, arr.ind = TRUE)) %>%
  mutate(number = row_number())

num_seats <- nrow(seat_coords)

seat_num <- function(i, j) {
  filter(seat_coords, row == i & col == j)$number
}

seat_coords
```

The only row/column information we need to retain is the visible neighbors.
Let's find the nearest seat in a direction:
```{r}
neighbor_seat <- function(i, j, di, dj) {
  while (((i <- i + di) %in% 1:nrow(seats)) &
    ((j <- j + dj) %in% 1:ncol(seats))) {
    if (seats[i, j]) {
      return(seat_num(i, j))
    }
  }
  NA
}
seat_coords[neighbor_seat(1, 1, 1, 1), ]
```
Then we can turn the counting of the neighbors into a matrix multiplication.
The sparse neighbor matrix, when multiplied with the seat vector,
returns a vector with the number of occupied neighbors for each seat.

```{r}
neighbors <- function(i, j) {
  n <- c(
    neighbor_seat(i, j, -1, -1),
    neighbor_seat(i, j, 0, -1),
    neighbor_seat(i, j, 1, -1),
    neighbor_seat(i, j, 1, 0),
    neighbor_seat(i, j, 1, 1),
    neighbor_seat(i, j, 0, 1),
    neighbor_seat(i, j, -1, 1),
    neighbor_seat(i, j, -1, 0)
  )
  n <- n[!is.na(n)]
  sparseMatrix(
    i = rep(seat_num(i, j), length(n)),
    j = n,
    dims = c(num_seats, num_seats)
  )
}
```
This bit is still loopy and it takes a couple of minutes to fill the matrix.
```{r}
neighbor_matrix <- sparseMatrix(c(), c(), dims = c(num_seats, num_seats))
for (number in 1:num_seats) {
  row <- seat_coords$row[[number]]
  col <- seat_coords$col[[number]]
  neighbor_matrix <- neighbor_matrix + neighbors(row, col)
}
dim(neighbor_matrix)
```
But iterating is a lightning fast vector operation now!
```{r}
life_step <- function(d) {
  pop <- (neighbor_matrix %*% d)[, 1]
  d <- (!d & pop == 0) | (d & pop < 5)
}
state <- rep(0, num_seats)
next_state <- life_step(state)
while (any(next_state != state)) {
  state <- next_state
  next_state <- life_step(state)
}
sum(state)
```
