---
title: "--- Day 10: Adapter Array ---"
author: Fleur Kelpin
date: Dec 10, 2020
output: github_document
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
input <- readr::read_csv("day10.txt", col_names = FALSE)[[1]] %>%
  prepend(0) %>%
  {append(., max(.) + 3)} %>%
  sort()
  
as_tibble(input)
```

# Part 1

Find a chain that uses all of your adapters to connect the charging outlet to
your device's built-in adapter and count the joltage differences between the
charging outlet, the adapters, and your device. **What is the number of 1-jolt
differences multiplied by the number of 3-jolt differences?**

```{r}
input %>%
  diff() %>%
  { sum(. == 1) * sum(. == 3) }
```

# Part 2

**What is the total number of distinct ways you can arrange the adapters to
connect the charging outlet to your device?**

Use dynamic programming to reuse the number of ways you can connect the lower
jolt adapters to the outlet to compute the number of ways for the next adapter.

```{r}
adapters <- tibble(jolt = input, ways=as.numeric(NA))
adapters[1, 'ways'] = 1

while((index <- detect_index(adapters$ways, is.na)) > 0 ){
  joltage <- adapters[index, 'jolt']
  prev_adapters <- filter(adapters, jolt %>% between(joltage-3, joltage-1))
  adapters[index, 'ways'] = sum(prev_adapters$ways)
}
adapters
options(digits = 15)
max(adapters$ways)
```
