---
title: "--- Day 17: Conway Cubes ---"
author: Fleur Kelpin
date: Dec 17, 2020
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
n <- 8
steps <- 6
input <- readLines("day17.txt") %>%
  str_split("") %>%
  unlist() %>%
  map_lgl(~ . == "#") %>%
  matrix(nrow = n, byrow = T)
input
```
# Part 1
Starting with your given initial configuration, simulate six cycles. How many
cubes are left in the active state after the sixth cycle?
```{r}
dim <- c(n + 2 * steps, n + 2 * steps, 1 + 2 * steps)
state <- array(0, dim = dim)
for(x in 1:n) {
  for(y in 1:n) {
    state[x+6,y+6,7] <- input[x,y]
  }
}
```
```{r}
neighbors <- function(pos) {
  result <- gtools::permutations(3, 3, c(0,-1,1), repeats.allowed = T) %>%
    matrix(ncol = 3, byrow = F)
  result[,1] <- result[,1] + pos[[1]]
  result[,2] <- result[,2] + pos[[2]]
  result[,3] <- result[,3] + pos[[3]]
  result[which(
    between(result[,1], 1, dim[1]) & 
     between(result[,2], 1, dim[2]) &
     between(result[,3], 1, dim[3])),]
}
```

```{r}
for(i in 1:6) {
  next_state <- array(0, dim = dim)
  for(x in 1:dim[1]) {
    for(y in 1:dim[2]) {
      for(z in 1:dim[3]) {
        pop <- sum(state[neighbors(c(x,y,z))]) - state[x,y,z]
        next_state[x,y,z] <- 
          (!state[x,y,z] & pop == 3) | 
          (state[x,y,z] & between(pop, 2, 3))
      }
    }
  }
  state <- next_state
}
sum(state)
```
# Part 2
For some reason, your simulated results don't match what the experimental energy
source engineers expected. Apparently, the pocket dimension actually has four
spatial dimensions, not three.
```{r}
dim <- c(n + 2 * steps, n + 2 * steps, 1 + 2 * steps, 1 + 2 * steps)
state <- array(0, dim = dim)
for(x in 1:n) {
  for(y in 1:n) {
    state[x+6,y+6,7,7] <- input[x,y]
  }
}
```
```{r}
neighbors <- function(pos) {
  result <- gtools::permutations(3, 4, c(0,-1,1), repeats.allowed = T) %>%
    matrix(ncol = 4, byrow = F)
  result[,1] <- result[,1] + pos[[1]]
  result[,2] <- result[,2] + pos[[2]]
  result[,3] <- result[,3] + pos[[3]]
  result[,4] <- result[,4] + pos[[4]]
  result[which(
    between(result[,1], 1, dim[[1]]) & 
     between(result[,2], 1, dim[[2]]) &
     between(result[,3], 1, dim[[3]]) &
     between(result[,4], 1, dim[[4]])),]
}
```

```{r}
for(i in 1:6) {
  next_state <- array(0, dim = dim)
  for(x in 1:dim[1]) {
    for(y in 1:dim[2]) {
      for(z in 1:dim[3]) {
        for(w in 1:dim[4]) {
          pop <- sum(state[neighbors(c(x,y,z,w))]) - state[x,y,z,w]
          next_state[x,y,z,w] <- 
            (!state[x,y,z,w] & pop == 3) | 
            (state[x,y,z,w] & between(pop, 2, 3))
        }
      }
    }
  }
  state <- next_state
}
sum(state)