---
title: "--- Day 22: Crab Combat ---"
author: Fleur Kelpin
date: Dec 22, 2020
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
input_player1 <- read_csv("day22-player1.txt", col_names = c("card"))$card
input_player2 <- read_csv("day22-player2.txt", col_names = c("card"))$card
input_player1
input_player2
```
# Part 1
> Before the game starts, split the cards so each player has their own deck
(your puzzle input). Then, the game consists of a series of rounds: both players
draw their top card, and the player with the higher-valued card wins the round.
The winner keeps both cards, placing them on the bottom of their own deck so
that the winner's card is above the other card. If this causes a player to have
all of the cards, they win, and the game ends.

> Once the game ends, you can calculate the winning player's score. The bottom
card in their deck is worth the value of the card multiplied by 1, the
second-from-the-bottom card is worth the value of the card multiplied by 2, and
so on. Play the small crab in a game of Combat using the two decks you just
dealt. What is the winning player's score?

```{r}
p1 <- input_player1
p2 <- input_player2
num_cards <- length(input_player1) + length(input_player2)
while (between(length(p1), 1, num_cards-1)) {
  #print(list(p1 = p1, p2 = p2))
  trick <- c(p1[[1]], p2[[1]])
  p1 <- tail(p1, n = -1)
  p2 <- tail(p2, n = -1)
  sorted <- sort(trick, decreasing = T)
  if (all(trick == sorted)) {
    p1 <- append(p1, sorted)
  } else {
    p2 <- append(p2, sorted)
  }
}
winning_hand <- c(p1,p2)
sum(winning_hand * (num_cards:1))
```