---
title: "--- Day 23: Crab Cups ---"
author: Fleur Kelpin
date: Dec 23, 2020
output: github_document
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
input <- "362981754" %>%
  str_split("") %>%
  unlist() %>%
  as.integer()
```
# Part 1
> Using your labeling, simulate 100 moves. What are the labels on the cups after
cup 1?

```{r}
get_destination <- function(current) {
  if (current == 1) {
    ncups
  } else {
    current - 1
  }
}
```
```{r}
cups <- input
ncups <- length(cups)
for(i in 1:100) {
  current <- cups[[1]]
  triple <- cups[2:4]
  rest <- cups[5:ncups]
  destination <- get_destination(current)
  while (!any(rest == destination)) {
    destination <- get_destination(destination)
  }
  index <- which(rest == destination)[[1]]
  if (index == length(rest)) {
    cups <- c(rest, triple, current)
  } else {
    cups <- c(rest[1:index], triple, rest[(index+1):length(rest)], current)
  }
}
cups
```
# Part 2

> Due to what you can only assume is a mistranslation (you're not exactly fluent
in Crab), you are quite surprised when the crab starts arranging many cups in a
circle on your raft - one million (1000000) in total.

> Your labeling is still correct for the first few cups; after that, the
remaining cups are just numbered in an increasing fashion starting from the
number after the highest number in your list and proceeding one by one until one
million is reached. (For example, if your labeling were 54321, the cups would be
numbered 5, 4, 3, 2, 1, and then start counting up from 6 until one million is
reached.) In this way, every number from one through one million is used exactly
once.

```{r}
cups <- 1:1000000
cups[1:length(input)] <- input
ncups <- length(cups)
```

>After discovering where you made the mistake in translating Crab Numbers, you
realize the small crab isn't going to do merely 100 moves; the crab is going to
do ten million (10000000) moves!

> Determine which two cups will end up immediately clockwise of cup 1. What do
you get if you multiply their labels together?

Use a data structure that can be updated in O(1) instead of keeping all cup
labels in order in the vector cause right now the entire vector needs to be
shuffled around for each iteration.

Store the label of the next cup (clockwise) in the vector instead.
```{r}
next_cup <- integer(ncups)
for(i in 1:(ncups - 1)) {
  next_cup[cups[i]] <- cups[[i+1]]
}
next_cup[cups[[ncups]]] <- cups[[1]]
current <- cups[[1]]
```
Then we can execute the desired number of moves more quickly.
```{r}
tictoc::tic()
for (i in 1:10000000) {
  destination <- get_destination(current)
  
  triple1 <- next_cup[[current]]
  triple2 <- next_cup[[triple1]]
  triple3 <- next_cup[[triple2]]
  while (destination %in% c(triple1, triple2, triple3)) {
    destination <- get_destination(destination)
  }
  
  next_current <- next_cup[[triple3]]
  next_cup[triple3] <- next_cup[[destination]]
  next_cup[destination] <- triple1
  next_cup[current] <- next_current
  current <- next_current
}
tictoc::toc()
prod(next_cup[[1]], next_cup[[next_cup[[1]]]])
```
