---
title: "--- Day 21: Allergen Assessment ---"
author: Fleur Kelpin
date: Dec 21, 2020
output: github_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
input <- tibble(line = read_lines("day21.txt")) %>%
  mutate(row = row_number()) %>%
  extract(line, c("ingredient", "allergen"), "(.*) \\(contains (.*)\\)") %>%
  separate_rows(allergen, sep = ", ") %>%
  separate_rows(ingredient, sep = " ")
input
```
# Part 1
>Each allergen is found in exactly one ingredient. Each ingredient contains zero
or one allergen. Allergens aren't always marked; when they're listed (as in
(contains nuts, shellfish) after an ingredients list), the ingredient that
contains each listed allergen will be somewhere in the corresponding ingredients
list. However, even if an allergen isn't listed, the ingredient that contains
that allergen could still be present: maybe they forgot to label it, or maybe it
was labeled in a language you don't know.

>Determine which ingredients cannot possibly contain any of the allergens in
your list. How many times do any of those ingredients appear?

This was a joy to write in R! (Once I figured out which methods to call.)
The input is already tidied. This method splits a data frame on row and then
joins them back together to find the ingredients that are listed in each of the
rows.
```{r}
shared_ingredients <- function(df) {
  ingredient <- split(df, df$row) %>%
    Reduce(function(df1, df2) inner_join(df1, df2, by = "ingredient"), .) %>%
    { .$ingredient }
  tibble(
    allergen = df$allergen[[1]],
    ingredient = ingredient
  )
}
shared_ingredients(filter(input, allergen=="fish"))
```
Then split the input on allergen and call the shared_ingredients method to find
the suspect ingredients for each allergen
```{r}
suspect_ingredients <- split(input, input$allergen) %>%
  map_df(shared_ingredients)
suspect_ingredients
```
Finally count the occurrences of the safe ingredients on rows
```{r}
input %>%
  filter(!ingredient %in% suspect_ingredients$ingredient) %>%
  group_by(ingredient) %>%
  summarize(row) %>%
  unique() %>%
  nrow()
```
# Part 2
>Now that you've isolated the inert ingredients, you should have enough
information to figure out which ingredient contains which allergen.

Keep filtering out the suspect ingredients that match exactly one allergen.
```{r}
dangerous_ingredients <-
  tibble(ingredient = character(), allergen = character())

while (nrow(suspect_ingredients) > 0) {
  suspect_ingredients <-
    suspect_ingredients %>%
    filter(!ingredient %in% dangerous_ingredients$ingredient) %>%
    filter(!allergen %in% dangerous_ingredients$allergen)
  discovered <- suspect_ingredients %>%
    group_by(allergen) %>%
    filter(length(ingredient) == 1)
  dangerous_ingredients <- union(dangerous_ingredients, discovered)
}

dangerous_ingredients
```
>Arrange the ingredients alphabetically by their allergen and separate them by 
commas to produce your canonical dangerous ingredient list.

```{r}
dangerous_ingredients %>%
  arrange(allergen) %>%
  { .$ingredient } %>%
  str_flatten(",")
```
